Title
========================================================

### Synopsis

### Loading, Processing and Formatting the Dataset

```{r cache=TRUE}
stormData <- read.csv('StormData.bz2')
```

Since we're going to be focused on analyzing the event type's (EVTYPE) impact against health and economic livelihood of the population, we can start by checking to make sure we have the columns properly named and typed.

```{r}
colnames(stormData)
```

Health impact would likely be measured best in number of fatalities and injuries (FATALITIES and INJURIES as the respective column names).

For the economic analysis we have PROPDMG and CROPDMG and their exp values in PROPDMGEXP and CROPDMGEXP

```{r}
unique(stormData$PROPDMGEXP)
```

Since the values have ranges from K, M, B to 1-9, we can assume K, M, B for thousand, million, billion and 1-9 as 10^n.  Now subset the columns we are taking to be relevant to the investigation, get rid of any zero values that we don't need.

```{r cache=TRUE}
st <- subset(stormData, FATALITIES > 0 | INJURIES > 0 | PROPDMG > 0 | CROPDMG > 0, select = c("STATE__", "BGN_DATE", "END_DATE", "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP"))
```

Let's total up the fatalities and injuries

```{r}
st$totalHealth <- st$FATALITIES + st$INJURIES
```

Now convert the dollar amounts for crop and property damage.

```{r}
st$propdamage <- st$PROPDMG
st$PROPDMGEXP <- as.character(st$PROPDMGEXP)
st$PROPDMGEXP[tolower(st$PROPDMGEXP) == "h"] <- "2"
st$PROPDMGEXP[tolower(st$PROPDMGEXP) == "k"] <- "3"
st$PROPDMGEXP[tolower(st$PROPDMGEXP) == "m"] <- "6"
st$PROPDMGEXP[tolower(st$PROPDMGEXP) == "b"] <- "9"
st$PROPDMGEXP[st$PROPDMGEXP == "" | st$PROPDMGEXP == "-" | st$PROPDMGEXP == "?" | st$PROPDMGEXP == "+"] <- "0"
st$PROPDMGEXP <- as.numeric(st$PROPDMGEXP)
st$propdamage <- (10^st$PROPDMGEXP) * st$PROPDMG
```

```{r}
st$cropdamage <- st$CROPDMG
st$CROPDMGEXP <- as.character(st$CROPDMGEXP)
st$CROPDMGEXP[tolower(st$CROPDMGEXP) == "h"] <- "2"
st$CROPDMGEXP[tolower(st$CROPDMGEXP) == "k"] <- "3"
st$CROPDMGEXP[tolower(st$CROPDMGEXP) == "m"] <- "6"
st$CROPDMGEXP[tolower(st$CROPDMGEXP) == "b"] <- "9"
st$CROPDMGEXP[st$CROPDMGEXP == "" | st$CROPDMGEXP == "-" | st$CROPDMGEXP == "?" | st$CROPDMGEXP == "+"] <- "0"
st$CROPDMGEXP <- as.numeric(st$CROPDMGEXP)
st$cropdamage <- (10^st$CROPDMGEXP) * st$CROPDMG
```

### Results

Now we can begin to look at the health impact by environmental disaster.  We can sum the total events by type, showing the top fifteen worst

```{r}
onlySums <- st[, c("EVTYPE", "FATALITIES", "INJURIES", "propdamage", "totalHealth")]
agged <- aggregate(. ~ EVTYPE, onlySums, sum)
head(agged[order(-agged$totalHealth), c("EVTYPE", "totalHealth")], 15)
```

Based on the total counts above, tornadoes have the worst impact on human life.

NOTE -> see if there is a correlation between health and economic risk.

